\Large Last Writer Slices and Communication Traps

### 2013-11-03 paper.tex
committing new text in intro, background, and formalism

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to the simplicity of
sequential reasoning, multi-threaded programs require complex reasoning about
many threads of control and their interactions.  In shared memory programs,
threads interact by reading and writing memory locations.  Threads' reads and
writes interleave arbitrarily and nondeterministically, with each interleaving
potentially yielding different execution behavior.  The complexity of reasoning
about thread interactions is a central challenge to building multi-threaded
software.

When programmers reason incorrectly, they permit some interleavings that should
be forbidden.  When such interleavings occur, program executions may fail,
experience performance degradation, or suffer other undesirable behavior.
Preventing such undesirable behavior is a fundamental challenge faced by
software engineers.  An important way to understand program behavior is to
monitor which read and write operations depend on one another.  Memory
operations are dependent if their order determines the values they read or
write.  The value of dependence information is made especially clear by its
prevalent use during manual program debugging and in support of program
analysis tools.

Debugging a concurrent program requires understanding how threads interact and
dependence information furnishes that understanding.  Programmers debug
programs more effectively when they can ask questions about the provenance of
values during a program's execution~\cite{slicingsurvey,whylineicse}.
Dependence information helps programmers with the difficult task of identifying
control- and data-flows that influenced a particular data value at a particular
point.  In concurrent programs, data provenance information is vital because
influential control- and data-flows may be in any thread, making the task of
understanding the origin of a value much more complex.

Compounding this complexity, many concurrency bugs manifest as failures rarely.
When a deployed system fails and a programmer receives a bug report,
reproducing the reported failure to collect dependence information may be
difficult or impossible.  It is essential to collect as much information as
possible in deployed systems to reduce the need for time-consuming forensic bug
reproduction.  Unfortunately, prior approaches to tracking dependences in
concurrent code~\cite{raceslicing,tipslicingsurvey} impose high run time
overheads.  These overheads are prohibitively high for use in production.

@@

Dependence information is also an important part of many program analysis
tools.  Prior work has used dependence information to reduce the burden of
making reliable, efficient concurrent software.  Bug
detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
analyze inter-thread dependences to find buggy behavior and help programmers
understand their errors.
Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
dependences to find optimization opportunities.  Dependences are also essential
to some program understanding systems~\cite{oshatr} and dynamic checkers for
thread interaction properties~\cite{oshajava,velodrome}.
Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
order of dependent events to help reproduce failures.

@@

These systems would be especially valuable, if run in production systems on
deployed code.  However, as with dependence tracking on its own, most of these
systems have overheads too high for use in production (5-100x
slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.

The key to reducing the overhead of dependence tracking is to carefully select
and monitor only a {\em subset} of dependences.  We start from two simple
observations: (1) Reads are frequent so we must minimize tracking overheads on
reads, and (2) In concurrency analyses, {\em inter-thread} dependences are most
important.  Based on these observations, we focus on dependences that reflect
{\em communication} between threads.  Communication occurs when a thread reads
or overwrites a memory location that was most recently written by a different
thread; {\em i.e.}, inter-thread read-after-write (RAW) and write-after-write
(WAW).  Tracking only communication dependences preserves enough information to
implement useful applications and analyses, but has low enough overhead for
always-on use in production systems.

Framed by these observations, we demonstrate in this work that a 
versatile form of inter-thread dependence-tracking analysis has inherent
overheads that are low enough for use in deployed systems.  There are two parts
to our work:

@@

\item{We discuss last-writer slices, an abstraction that dynamically tracks
memory updates, capturing inter-thread dependence information.  We
design and implement system support for collecting last writer slices.  We
present two cases studies that show that last writer slices are useful for
diagnosing real-world bugs.}

\item{Building on last-writer slices, we describe and implement \ctraps, which
exposes inter-thread communication through an extensible trap interface.
We also look at an optimization to \ctraps that explores trading off
precision for performance.}

\item{We demonstrate that \ctraps provides a simple platform for building
dependence analyses by implementing two analyses from prior
work~\cite{cci,defuse,recon}. }

\item{We implement last writer slicing and \ctraps and evaluate their performance using a
variety of important applications ({\em e.g.}, MySQL, memcached and the PARSEC
benchmarks~\cite{parsec}).  We show that in many cases our designs have overheads
low enough for use in deployment (0--15\%).}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.6]{figs/BasicLWT.pdf}
\caption{\label{fig:basicLWT}Basic operation of the last writer table. }
\end{figure}

@@

\subsubsection{Escape Analysis}
We use a local function escape analysis to identify which accesses are to
memory locations that can be statically guaranteed to be inaccessible to code
running in other threads.  When a thread accesses a location that is
inaccessible to code in other threads, there is no way that access can be a
communicating access.  Such accesses are not instrumented by our compiler.

We rely on {\tt GCC}'s built-in escape analysis to determine if a memory
operation might lead to communication.  This analysis is a function escape
analysis.  Locations determined not to escape function scope also cannot escape
the thread executing the function.  However, using an inter-procedural thread
escape analysis may eliminate more instrumentation sites.  Investigating better
instrumentation strategies is orthogonal, but an interesting subject for future
work.

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_i = (p^{i}_{1},
p^{i}_{2}, \ldots, p^{i}_{k_{i}})$ is the sequence of $k_{i}$ dynamic
instructions executed by thread $i$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = (q_{0}, \ldots, q_{e})$ where each $q \in Q$ is a
$p^{i}_{j}$ from one of the $k_i$ instructions of thread $i$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
our \lwt implementation, must be the same as $LWS[m]$ at all points in an
execution.  Correctness in the \lwt requires {\em soundness} and {\em
completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is
written.  A {\em complete} design always updates $LWT[m]$ when $m$ is written.
Our compiler inserts \lwt updates before every write
operation (for completeness) and only before write operations (for soundness).
Formally, given a program, $P$, our compiler pass transforms it to a new
program, $P'$.  Each thread in $P'$ executes a modified sequence of
instructions: $P'_{i} \in P' = \{ I(p^{i}_{1}), p^{i}_{1}, \ldots,
I(p^{i}_{k_{i}}), p^{i}_{k_{i}} \}$.  $I(p^{t}_{j})$ sets $LWT[m] =
(t,p^{t}_{j})$ if and only if $p^{t}_{j} = Write(m)$ and is a no-op otherwise.

Thus, our \lwt is correct under the two main assumptions of our formalism: (1)
there are no data races in the execution; and (2) each memory location
referenced by a write refers to a unique region of memory that does not overlap
with another referenced region.

\paragraph{Performance, Atomicity, and Sequential Consistency}

We do not enforce the atomicity of \lwt updates and their corresponding program
accesses.  This design choice, made in prior work~\cite{fasttrack,recon},
allows us to omit the synchronization operations that enforce the atomicity of
these operations.  The benefit of omitting these synchronization operations is
that overhead of our analysis is reduced. The downside is that our analysis is
vulnerable to violations of the atomicity of \lwt updates and their
corresponding accesses. Importantly, however, these violations can only occur
in programs with data-races and all programs with data-races do not necessarily
corrupt the \lwt.

We use our \lwt formalism from above to characterize how data-races can
interfere with \lwt updates.  The relation $p \prec q$ denotes that $p$ happens
before $q$.  In the absence of data-races, any write by thread $t$ to $m$,
$p^{t}_{j} = Write(m)$, is ordered by program synchronization with $p^{t'}_{j'}
= Op(m)$, a later access to $m$ by any other thread $t'$ -- {\em i.e.},
$p^{t}_{j} \prec p^{t'}_{j'}$.  Our compiler inserts an \lwt update
$I(p^{t}_{j})$ into $P'$ before its write operation, $p^{t}_{j}$, so
$I(p^{t}_{j}) \prec p^{t}_{j}$.  Happens-before is transitive, so without
data-races, $I(p^{t}_{j}) \prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec
p^{t'}_{j'}$, meaning $I(p^{t}_{j})$ happens before all accesses that
$p^{t}_{j}$ happens before {\em and before their respective \lwt updates},
$I(p^{t'}_{j'})$.  Symmetrically, for any access $p^{t''}_{j''}$ in another
thread $t''$ that precedes $p^{t}_{j}$, $p^{t''}_{j''} \prec p^{t}_{j}$.  Our
compiler inserts each $I(p^{t}_{j})$ immediately before each $p^{t}_{j}$, so
$I(p^{t''}_{j''}) \prec p^{t''}_{j''} \prec I(p^{t}_{j}) \prec p^{t}_{j}$,
happens-before ordering $I(p^{t}_{j})$ after all preceding accesses to $m$ in
other threads, {\em and after their repective \lwt updates}.  Thus, at all
points $Q$ in a data-race free execution, given $q_r = I(p^{t}_{j})$, $q_p =
p^{t}_{j}$, and $q_r \prec q_p$, $\forall_{ \{q_s | q_r \prec q_s \prec q_p
\}}: q_s \ne Op(m) \wedge q_s \ne I(Op(m)) $.  In other words, \lwt updates and
their accesses are atomic.

With data-races, \lwt updates and their accesses may not be atomic ({\em i.e.},
such a $q_s$ can exist).  Instrumentation calls for racy accesses are not
ordered by happens-before, so they may interleave arbitrarily.  With races, an
execution could reach a state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}),
p^{t'}_{j'}, p^{t}_{j}, \ldots)$.  In such a state, $LWS[m] = (t,p^{t}_{j})$,
but $LWT[m] = (t',p^{t'}_{j'})$.  The \lwt was effectively not updated when
$p^{t}_{j}$ executed (violating completeness), or was effectively updated to
reflect that $p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was
(violating soundness).  This edge case is incorrect. However, we expect it is
also uncommon in practice, so last writer slices are usually correct, even in
racy programs.

>>>>

\section{Introduction}
Multi-threaded programming is challenging.  In contrast to the simplicity of
sequential reasoning, multi-threaded programs require complex reasoning about
many threads of control and their interactions.  In shared memory programs,
threads interact by reading and writing memory locations.  Threads' reads and
writes interleave arbitrarily and nondeterministically and each interleaving
can yield different behavior.  Reasoning about thread interactions is a central
challenge to building multi-threaded software because some thread interactions
have undesirable consequences, like causing a crash or degrading performance.

A common approach to this challenge is to reason about dependent memory
operations.  Memory operations are dependent if their order determines the
values they read or write.  The utility of dependence information is made
especially clear by its prevalent use during manual program debugging and in
support of program analysis tools.

Debugging a concurrent program requires understanding how threads interact and
dependence information furnishes that understanding.  Dependence information
helps programmers with the difficult task of identifying control- and
data-flows that influenced a particular data value at a particular point.
Programmers debug programs more effectively when they can ask questions about
the provenance of values during a program's
execution~\cite{slicingsurvey,whylineicse}.  In concurrent programs, data
provenance information is particularly valuable because data may originate from
any of a potentially large number of threads, making reasoning complex.

Compounding this complexity, many concurrency bugs manifest as failures rarely.
When a deployed system fails and a programmer receives a bug report,
reproducing the reported failure to collect provenance information may be
difficult or impossible.  It is essential to collect as much information as
performance margins permit from deployed systems, to reduce the need for
time-consuming failure reproduction.  Unfortunately, prior approaches to
tracking dependences in concurrent code~\cite{raceslicing,tipslicingsurvey}
impose high run time overheads.  These overheads are prohibitively high for use
in production.

@@

Dependence information is also an important part of many program analysis
tools.    Bug detection~\cite{avio,fasttrack,raceslicing,dmtracker} and
debugging~\cite{tipslicingsurvey,bugaboo,recon,cci,defuse,conseq,falcon} tools
analyze inter-thread dependences to find buggy behavior and help programmers
understand their errors.
Profilers~\cite{threadclustering,schedpredictionmodel} analyze inter-thread
dependences to find optimization opportunities.  Dependences are also essential
to some program understanding systems~\cite{oshatr} and dynamic checkers for
thread interaction properties~\cite{oshajava,velodrome}.
Record-Replay~\cite{chimera,doubleplay,fdr,rtr} systems track and record the
order of dependent events to help reproduce failures.  These systems would be
even more valuable, if their underlying dependence analysis was run in
production.  However, the heavy-weight analyses proposed by many of these
systems impose overheads too high for production use (5-100x
slowdown)~\cite{raceslicing,dmtracker,velodrome,recon,defuse,oshajava,chimera,coredet,stm}.

@@

Our goal in this work is to collect, with deployable efficiency, useful
dependence information that captures data provenance for use in debugging and
serves as a foundation for more sophisticated program analyses.  We start from
two simple observations: (1) Reads are frequent so we must minimize tracking
overheads on reads, and (2) In concurrency analyses, {\em inter-thread}
dependences are most important.  Based on these observations, we focus first on
tracking write-after-write dependences (WAWs), which reflect data provenance.
Building on that, we describe support for tracking {\em communication
dependences}.  Communication occurs when a thread reads or overwrites a memory
location that was most recently written by a different thread; {\em i.e.},
inter-thread read-after-write dependences (RAWs) as well as (WAWs).  In this
work, we demonstrate that this form of light-weight dependence tracking is
useful and versatile, and has inherent overheads that are low enough for use in
deployed systems.  There are two parts to our work:

@@

\item{We describe last writer slicing, a dynamic analysis that tracks
memory updates, capturing data provenance information.  We
design and implement system support for collecting last writer slices.  }

\item{Building on last writer slicing, we describe and implement \ctraps, which
exposes inter-thread communication through an extensible trap interface.  }

\item{We evaluate the utility of last writer slices and \ctraps. We first
present several case studies that show that last writer slices are useful for
debugging real-world bugs.  We then show how \ctraps is useful for building
dependence analyses by implementing two analyses from prior
work~\cite{cci,defuse,recon}. }

\item{We evaluate our system's performance using a variety of important
applications ({\em e.g.}, MySQL, memcached and the PARSEC
benchmarks~\cite{parsec}).  We show that in many cases our designs have
overheads low enough for use in deployment (0--15\%).}

@@

\begin{figure}[h]
\centering
\includegraphics[scale=.6]{figs/BasicLWT.pdf}
\caption{\label{fig:basicLWT}Basic operation of the last writer table. Different shapes signify different distinct memory operations. }
\end{figure}

@@

\subsubsection{Escape Analysis}
We use a local function escape analysis to identify which accesses are to
memory locations that can be statically guaranteed to be inaccessible to code
running in other threads.  When a thread accesses a location that is
inaccessible to code in other threads, there is no way that access can be a
communicating access.  Such accesses are not instrumented by our compiler.  
We rely on {\tt GCC}'s built-in function escape analysis to determine if a
memory operation might lead to communication.  Locations determined not to
escape function scope also cannot escape the thread executing the function.
%Using an inter-procedural thread escape analysis may eliminate more
%instrumentation sites.

\subsection{\lwt Formalism and Correctness}
\label{sec:lwssoundness}
To discuss the correctness of our \lwt-based last writer slice support, we
first develop a formal description of last writer slices.  We consider an
N-threaded program $P = \{P_1, P_2, \ldots P_N\}$, where $P_t = (p^{t}_{1},
p^{t}_{2}, \ldots, p^{t}_{k_{t}})$ is the sequence of $k_{t}$ dynamic
instructions executed by thread $t$.  Without loss of generality, we assume all
instructions are memory reads and writes, with the form $Op(m)$ where $Op$ is
$Read$ or $Write$ and $m$ is a uniquely addressed memory location.
A point in an execution of $P$ after $e$ instructions have
executed is defined by $Q = \{q_{0}, \ldots, q_{e}\}$ where each $q \in Q$ is a
$p^{t}_{j}$ from one of the $k_t$ instructions of thread $t$.  We introduce
$LWS[m]$ -- the Last Writer Slice for memory location $m$.   At any point $Q$,
$LWS[m]$ holds $(t,p^{t}_{\ell w})$, where $p^{t}_{\ell w} = max_{\ell}( 
q_{\ell} = Write(m) \in Q)$ and $t$ is the thread that
executed $p^{t}_{\ell w}$. Informally, $p^{t}_{\ell w}$ is the last instruction
in $Q$ to write $m$ and $t$ is the thread that executed it.  For a
never-written location $n$, $LWS[n] = (\emptyset,\emptyset)$ indicating it is
unwritten.

For our design to be correct, $LWT[m]$, the entry for a memory location $m$ in
our \lwt implementation, must be the same as $LWS[m]$ at all points in an
execution.  Correctness in the \lwt requires {\em soundness} and {\em
completeness}.  A {\em sound} design only updates $LWT[m]$ when $m$ is written.
A {\em complete} design always updates $LWT[m]$ when $m$ is written.  Our
compiler inserts \lwt updates before every write operation (for completeness)
and only before write operations (for soundness).  Formally, given a program,
$P$, our compiler pass transforms it to a new program, $P'$.  Each thread in
$P'$ executes a modified sequence of instructions: $P'_{t} \in P' = \{
I(p^{t}_{1}), p^{t}_{1}, \ldots, I(p^{t}_{k_{t}}), p^{t}_{k_{t}} \}$.
$I(p^{t}_{j})$ sets $LWT[m] = (t,p^{t}_{j})$ if and only if $p^{t}_{j} =
Write(m)$ and is a no-op otherwise.  This reasoning shows our \lwt is both
sound, complete, and thus correct.

Next we discuss two nuances of our design: (1) the interaction of our \lwt
implementation with code containing data-races; and (2) the correctness of the
\lwt when a program addresses data that overlap in memory.

\paragraph{Consistency and Atomicity}

Our implementation of the \lwt does not add any synchronization to the program.
As a result our analysis does not incur the overheads or possible thread
schedule perturbation synchronization can cause.  However, there are two
potential issues that stem from this design choice. First, we do not enforce
memory consistency of meta-data reads and writes in different threads.  Second,
our implementation does not enforce the atomicity of \lwt updates and their
corresponding program accesses.  In the absence of data-races our \lwt
implementation is {\em always correct} and neither of these issues exist.  Both
atomicity and ordering is enforced by the program synchronization that ensures
consistency for the data being written by the instructions our system
instruments.

We use our \lwt formalism to show that without data-races, \lwt updates and
their corresponding memory accesses are atomic and well-ordered.  The relation
$p \prec q$ denotes that $p$ happens before $q$.  In correctly synchronized
code, any write by thread $t$ to $m$, $p^{t}_{j} = Write(m)$, is ordered by
program synchronization with $p^{t'}_{j'} = Op(m)$, a later access to $m$ by
any other thread $t'$ -- {\em i.e.}, $p^{t}_{j} \prec p^{t'}_{j'}$.  Our
compiler inserts an \lwt update $I(p^{t}_{j})$ into $P'$ before its write
operation, $p^{t}_{j}$. Thus, $I(p^{t}_{j}) \prec p^{t}_{j}$ due to program
order.  Happens-before is transitive, so without data-races, $I(p^{t}_{j})
\prec p^{t}_{j} \prec I(p^{t'}_{j'}) \prec p^{t'}_{j'}$, meaning $I(p^{t}_{j})$
happens before all accesses that $p^{t}_{j}$ happens before {\em and before
their respective \lwt updates}, $I(p^{t'}_{j'})$.  Symmetrically, for any
access $p^{t''}_{j''}$ in another thread $t''$ that precedes $p^{t}_{j}$,
$p^{t''}_{j''} \prec p^{t}_{j}$.  Our compiler inserts each $I(p^{t}_{j})$
immediately before each $p^{t}_{j}$, so $I(p^{t''}_{j''}) \prec p^{t''}_{j''}
\prec I(p^{t}_{j}) \prec p^{t}_{j}$, happens-before ordering $I(p^{t}_{j})$
after all preceding accesses to $m$ in other threads, {\em and after their
repective \lwt updates}.  Thus, at all points $Q$ in a data-race free
execution, given $q_r = I(p^{t}_{j})$, $q_p = p^{t}_{j}$, and $q_r \prec q_p$,
$\forall_{ \{q_s | q_r \prec q_s \prec q_p \}}: q_s \ne Op(m) \wedge q_s \ne
I(Op(m)) $.  In other words, \lwt updates and their accesses are atomic.

In a program with a data-race, there are some accesses, $p^{t}_{j}$ and
$p^{t'}_{j'}$, that are concurrent.  The \lwt updates for these accesses are
also concurrent because they are not transitively ordered by the accesses' ordering.  The unordered accesses and \lwt updates can interleave arbitrarily
because they are concurrent.  Some racy thread schedules may violate the
atomicity of accesses and their \lwt updates.  Such an execution could reach a
state $Q_{bad} = (\ldots, I(p^{t}_{j}), I(p^{t'}_{j'}), p^{t'}_{j'}, p^{t}_{j},
\ldots)$ and in that state, $LWS[m] = (t,p^{t}_{j})$, but $LWT[m] =
(t',p^{t'}_{j'})$.  The \lwt was effectively not updated when $p^{t}_{j}$
executed (violating completeness), or was effectively updated to reflect that
$p^{t'}_{j'}$ was $m$'s last writer, when $p^{t}_{j}$ was (violating
soundness).

According to the C++11 language specification, in the presence of a data-race,
a program's behavior is undefined and arbitrary state corruption can occur.  In
the presence of data-races, {\em any} analysis, including our \lwt analysis,
running on such a program could experience arbitrary corruption.  In practice
({\em e.g.}, on x86) these atomicity violations may cause limited \lwt
corruption, but even this is unlikely.  These atomicity violations only occur
only when accesses to the same data occur within a very short time interval,
which is uncommon, {\em and} an unserializable interleaving of accesses and
\lwt updates also occurs.

{\bf TODO: Figure?}
